# Epidemiology-the basics {#basics}

This chapter contains the basic principles of epidemiological analysis and how estimates of the risks associated with exposures can be obtained. From this chapter, the reader will have gained an understanding of the following topics:

- Methods for expressing risk and their use with different types of epidemiological study.
- Calculating risks based on calculations of the expected number of health counts in an area, allowing for the age--sex structure of the underlying population.
- The use of generalised linear models (GLMS) to model counts of disease and case--control indicators.
- Modelling the effect of exposures on health and allowing for the possible effects of covariates.
- Cumulative exposures to environmental hazards.


## Example 2.9: Estimating the SMR using a Poisson GLM {-}


```{r Ex 2.9 smr poisson one location}

# Finding MLE and SE of log(SMR) = beta0 on one single area
y <- 29  # Total observed death
E <- 19.88 # Expected deaths

summary(glm(y ~ offset(log(E)), family = "poisson"))

```
```{r Ex 2.9 smr poisson multiple locations, eval = FALSE}
# Finding MLE and SE of log(SMR) = beta0 over multiple areas
summary(glm(Y ~ offset(log(E)),family="poisson", data=data))

```


## Example 2.10: Estimating the SMR using quasi-likelihood {-}

```{r Ex 2.10 smr quasi-likelihood, eval = FALSE}
# Using quasi-likelihood to find the MLE and standard error of log(SMR) = beta0
summary(glm(y ~ offset(log(E)), family="quasipoisson"), data = data)

```


## Example 2.11: Modelling differences in SMRs in relation to differences in exposures {-}

```{r Ex 2.11 relative risk, eval = FALSE}
# Fitting a model to estimate the relative risk associated with air pollution
summary(glm(Y ~ offset(log(E)) + X1, family="poisson", data=data))

# Fitting a model to estimate the relative risk associated with air pollution 
# using a Quasi-Poisson approach
summary(glm(Y ~ offset(log(E)) + X1, family = "quasipoisson", data = data))

#Fitting a Poisson GLM with air pollution and deprivation
summary(glm(Y ~ offset(log(E)) + X1 + X2, family = "poisson", data = data))

# Fitting a Quasi-Poisson GLM with air pollution and deprivation
summary(glm(Y ~ offset(log(E)) + X1 + X2, family = "quasipoisson", data =
              data))

```

Perform tests between the deviances of two models.

```{r Ex 2.11 tests, eval = FALSE}

# Test 1: Effect on Quasi-Poisson models with and without deprivation
anova(
  glm(Y ~ offset(log(E)) + X1, family = "quasipoisson", data = data),
  # Model 1
  glm(Y ~ offset(log(E)) + X1 + X2, family = "quasipoisson", data = data),
  # Model2
  test = "Chisq"
) # Chi-Squared test

# Test 2: Effect on Quasi-Poisson models with and without air pollution
anova(
  glm(Y ~ offset(log(E)) + X1 + X2, family = "quasipoisson", data = data),
  # Model 1
  glm(Y ~ offset(log(E)) + X2, family = "quasipoisson", data = data),
  # Model 2
  test = "Chisq"
) # Chi-Squared test

```

## Example 2.12: Modelling the risks associated with lagged effects of air pollution {-}

```{r Ex 2.12 lagged effects, eval = FALSE}
# Fitting quasi-poisson model
glm(
  formula = Y ~ offset(log(E)) + X1 + X1t1 + X1t2,
  family  = "quasipoisson",
  data    = data
)
```

## Example 2.13: Estimating the odds ratio in a case-control study using a logistic model {-}  

```{r Ex 2.13 odds ratio, eval = FALSE}
# Fitting Odds Ratio
glm(formula = Y ~ 1,
    family = "binomial",
    data = data)
```

## Example 2.14: Estimating the odds ratio of asthma associated with proximity to roads {-}

```{r Ex 2.14 odds ratio asma, eval = FALSE}
# Fitting Odds Ratio
glm(Y ~ X, family = "binomial", data = data)

```



