# Spatial patterns in disease {#Spatial}

## Example 8.1: Empirical Bayes and Bayes smoothing of COPD mortality for 2010 {-}

Following [Example-5.2] we look back at the hospital admission rates for COPD, in England for 2010. We can implement the same model we used on [Example-5.2] in `nimble`.

```{r Ex 8.1 load data, message=FALSE, echo = FALSE, warning=FALSE, message= FALSE}

# Load nimble
library(nimble)
library(sf)

# Load data
# Reading in borders
england <- read_sf("data/englandlocalauthority.shp")

# Reading in data
observed <-
  read.csv(file = "data/copdmortalityobserved.csv", row.names = 1)

expected <-
  read.csv(file = "data/copdmortalityexpected.csv", row.names = 1)

```


### Nimble {-}

The following is the code for the Poisson-Gamma model implemented in `Nimble`.

```{r Ex 8.1 nimble model, results='hide', warning=FALSE, message= FALSE}

Example8_1Code <- nimbleCode({
  for (i in 1:N) {
    Y[i] ~ dpois(mu[i])
    # REVIEW: There is an intercept in the book, 
    # but then we wouldn't be able to compare it to example 5.2
    mu[i] <- E[i] * theta[i]
    # REVIEW: Same as before, the example in the book has theta[i] ~ Ga(a,a)
    theta[i] ~ dgamma(a, b)
    Y.fit[i] ~ dpois(mu[i])
  }
  
  # Priors
  a ~ dexp(lambda_a)
  b ~ dexp(lambda_b)

})

```

Define the constants, data and initials lists for the `Nimble` model.

```{r Ex 8.1 nimble set constants and inits}

# observations
y <- observed$Y2010
# offset
E <- expected$E2010
N <- length(y)
# Parameter of exponential prior for a
lambda_a <- 1
# Parameter of exponential prior for b
lambda_b <- 1

constants <- list(
  N = N,
  E = E,
  lambda_a = lambda_a,
  lambda_b = lambda_b
)
ex.data <- list(Y = y)

# Function to generate initial values
inits <-
  function()
    list(
      theta = rgamma(N, 1, 1),
      a = rexp(1, lambda_a),
      b = rexp(1, lambda_b),
      Y.fit = rpois(N, E)
    )

```

Define the parameter monitors and run the model.

```{r Ex 8.1 nimble run, error = FALSE, message = FALSE, results='hide'}

# parameters to monitor
params <- c("theta", "a", "b", "Y.fit")

mcmc.out <- nimbleMCMC(
  code = Example8_1Code,
  constants = constants,
  data = ex.data,
  # provide the combined data & constants as constants
  inits = inits,
  monitors = params,
  niter = 50000,
  nburnin = 20000,
  thin = 30,
  WAIC = TRUE,
  nchains = 2,
  summary = TRUE,
  samplesAsCodaMCMC = TRUE
)


```


Check the WAIC  and ESS.

```{r Ex 8.1 nimble WAIC and ESS}
print("WAIC")
mcmc.out$WAIC

print("min ESS")
min(coda::effectiveSize(mcmc.out$samples))


```

Show the trace plots and posterior summaries for each of the parameters.

```{r Ex 8.1 nimble traceplots}

mvSamples <- mcmc.out$samples

#trace plots of a
plot(mvSamples[, c("a")])

#trace plots of b
plot(mvSamples[, c("b")])

summary(mvSamples[, c("a", "b")])

#trace plots of theta's
for (i in 1:3)
  plot(mvSamples[, c(paste("theta[", i, "]", sep = ""))])

```

Now that we have checked the convergence of the chains we can plot the posterior mean and 95% CIs for each of the parameters.


```{r Ex 8.1 nimble posterior summary}

# Posterior summaries of theta_i
post_summary <- mcmc.out$summary$all.chains  %>% as.data.frame() %>%
  tibble::rownames_to_column("variable")

# Plot the mean and 95% CIs for the thetas
post_theta <-  post_summary[grepl("theta\\[", post_summary$variable), ]

par(mfrow = c(1, 1))
plot(
  post_theta$Mean,
  pch = 19,
  cex = 0.8,
  bty = "n",
  xlab = "Borough",
  ylab = "Posterior Summary Rate",
  ylim = c(min(post_theta$`95%CI_low`), max(post_theta$`95%CI_upp`))
)
for (i in 1:N)
  segments(i, post_theta$`95%CI_low`[i], i, post_theta$`95%CI_upp`[i])
abline(h = 1, lwd = 2, lty = 2)

# Posterior summary of fitted values
post_fitted <-  post_summary[grepl("Y.fit\\[", post_summary$variable), ]

# Plot mean and 95% CIs for the fitted values
par(mfrow = c(1, 1))
plot(
  y,
  post_fitted$Mean,
  ylim = c(min(post_fitted$`95%CI_low`), max(post_fitted$`95%CI_upp`)),
  xlab = "Observed",
  ylab = "Fitted",
  pch = 19,
  cex = 0.7,
  bty = "n"
)
for (i in 1:N)
  segments(y[i], post_fitted$`95%CI_low`[i], y[i], post_fitted$`95%CI_upp`[i])
abline(a = 0, b = 1)

```

### Stan

Load `stan` with options

```{r Ex 8.1 stan load, message=FALSE, echo = FALSE, warning=FALSE, message= FALSE}

rm(list=ls())

#loading rstan
library(rstan) 
library(loo) # To calculate WAIC and loo later
options(mc.cores = parallel::detectCores())
rstan_options(auto_write = TRUE)

# Load data
# Reading in borders
england <- read_sf("data/englandlocalauthority.shp")

# Reading in data
observed <-
  read.csv(file = "data/copdmortalityobserved.csv", row.names = 1)

expected <-
  read.csv(file = "data/copdmortalityexpected.csv", row.names = 1)


```

Write the `stan` model. This model is in a separate file called `Example8_1.stan` that will be called later.

```{r Ex 8.1 stan model, echo = FALSE}
system("cat functions/Example8_1.stan")
```


Define the data for the model, similar to `nimble`. 

```{r Ex 8.1 stan define data, error = FALSE, message = FALSE, results='hide'}
# observations
y <- observed$Y2010
# offset
E <- expected$E2010
N <- length(y)

# Define the data list 
ex.data <- list(
  N = length(y),
  Y = y,
  E = E,
  lambda_a = 1,
  lambda_b = 1
)

# Run the model in Stan

mod0 <- stan(
  file = "functions/Example8_1.stan",
  data = ex.data,
  chains = 3,
  iter = 10000,
  warmup = 3000,
  thin = 14,
  control = list(adapt_delta = 0.8, max_treedepth = 15),
  init = "random",
  pars = c("a", "b", "theta", "log_lik", "yfit"),
  include = TRUE
)

```

Compute the WAIC.

```{r Ex 8.1 stan WAIC, message= FALSE, warning=FALSE}

#computing WAIC using the package loo
loglik0 <- extract_log_lik(mod0)
waic0 <- waic(loglik0)
waic0

```


Show the trace plots and posterior summaries of the parameters.

```{r Ex 8.1 stan traceplots}

#traceplots of the vector of coefficients beta
traceplot(mod0,pars=c("a","b"))

traceplot(mod0,pars=c("theta[1]", "theta[2]", "theta[3]"))
```
```{r Ex 8.1 stan posterior summary}

summary_theta <-
  summary(mod0, pars = c("theta"), probs = c(0.05, 0.95))$summary %>% as.data.frame()


par(mfrow = c(1, 1))
plot(
  summary_theta$mean,
  pch = 19,
  cex = 0.8,
  bty = "n",
  xlab = "Borough",
  ylab = "Posterior Summary Rate",
  ylim = c(min(summary_theta$`5%`), max(summary_theta$`95%`))
)
for (i in 1:N)
  segments(i, summary_theta$`5%`[i], i, summary_theta$`95%`[i])
abline(h = 1, lwd = 2, lty = 2)

# Posterior summary of fitted values

summary_fit <-
  summary(mod0, pars = c("yfit"), probs = c(0.05, 0.95))$summary %>% as.data.frame()


# Plot mean and 95% CIs for the fitted values
par(mfrow = c(1, 1))
plot(
  y,
  summary_fit$mean,
  ylim = c(min(summary_fit$`5%`), max(summary_fit$`95%`)),
  xlab = "Observed",
  ylab = "Fitted",
  pch = 19,
  cex = 0.7,
  bty = "n"
)
for (i in 1:N)
  segments(y[i], summary_fit$`5%`[i], y[i], summary_fit$`95%`[i])
abline(a = 0, b = 1)
```

# Example 8.3: Fitting a conditional spatial model in NIMBLE

## Nimble



