# But can the data be trusted {#trusted}

This chapter considers some of the issues that will arise when dealing with 'real data’. Data will commonly have missing
values and may be measured with error. This error might be random or may be due to systematic patterns in how it was
collected. From this chapter, the reader will have gained an understanding of the following topics:

- Classification of missing values into missing at random or not at random.
- Methods for imputing missing values.
- Various measurement models including classical and Berkson.
- The attenuation of regression coefficients under measurement error.
- Preferential sampling, where the process that determines the locations of monitoring sites and the process being modelled are in some ways dependent.
- How preferential sampling can bias the measurements that arise from environmental monitoring networks

## Solutions to Selected Exercises

### PLACEHOLDER: Repeat the Black Smoke analysis conducted in Watson et al. 2019. in the "naive" model. 

If INLA is not already installed, it needs to be done manually (https://www.r-inla.org/download-install):

```{r install inla, eval=FALSE}
install.packages(
  "INLA",
  repos = c(getOption("repos"), INLA = "https://inla.r-inla-download.org/R/stable"),
  dep = TRUE
)
```

After INLA is installed load all the necessary packages.

```{r Ex load packages , eval = FALSE}
library(INLA)
library(mvtnorm)
library(boot)
library(geoR)
library(reshape2)
library(sp)
library(ggplot2)
```

Assuming the data is held in the file `BlackSmokePrefData.csv` in a `/data` directory:

```{r data clean and standardize, eval = FALSE}

BlackSmokePrefData <- read.csv("data/BlackSmokePrefData.csv")

head(BlackSmokePrefData)

## standardize location coordinates

sd_x <- sd(BlackSmokePrefData[, c(2)])
sd_y <- sd(BlackSmokePrefData[, c(3)])

BlackSmokePrefData[, 2] <- BlackSmokePrefData[, 2] / sd_x
BlackSmokePrefData[, 3] <- BlackSmokePrefData[, 3] / sd_y 
```

We need to reshape the data to have one observation per row, as required by INLA.

```{r reshape, eval = FALSE}
yearmeans = colMeans(BlackSmokePrefData[,-c(1, 2, 3)], na.rm = T)  ## save means for each year

BlackSmokePrefData <-
  melt(
    BlackSmokePrefData,
    id.vars = c(1, 2, 3),
    variable.name = 'year',
    value.name = 'bsmoke'
  )
BlackSmokePrefData$year = as.numeric(as.character(factor(BlackSmokePrefData$year, labels =
                                                           66:96)))
## standardize by mean of mean of each year to make it unitless
BlackSmokePrefData$bsmoke = BlackSmokePrefData$bsmoke / mean(yearmeans)
## log-transform to eliminate right skew
BlackSmokePrefData$bsmoke = log(BlackSmokePrefData$bsmoke)
```

Producing the 2-d mesh of the UK:

```{r mesh analysis, eval = FALSE}
no_sites = as.numeric(length(unique(BlackSmokePrefData$site))) # number of sites
no_T = as.numeric(length(unique(BlackSmokePrefData$year))) # number of years

ncol = 100 # grid for projection
nrow = 100
L = nrow * ncol # number of grid sites

# Form the regular mesh #
# Create a rough convex boundary for the UK #
# Form the grid independently from the sites to avoid preferential grid selection #
UK_domain = cbind(c(2, 7.7, 7.7, 6, 4, 1), c(0.5, 0.5, 6, 13.5, 13.5, 12))
hull = inla.nonconvex.hull(cbind(BlackSmokePrefData$east, BlackSmokePrefData$north))

cutoff_dist = 16000 / sd_x # 16km as min edge
max.edge = 100000 / sd_x # 100km max edge as in Shaddick and Zidek

# NOTE: JOHNNY, I HAD TO COMMENT THE CUTOFF VARIABLE CAUSE OTHERWISE IT WOULDN'T WORK

mesh = inla.mesh.2d(
  loc = cbind(BlackSmokePrefData$east, BlackSmokePrefData$north),
  boundary = hull,
   offset = c(0.1, 0.2),
   max.edge = c(cutoff_dist, 0.5),
   #cutoff = c(cutoff_dist, 1), 
   min.angle = 26
)

plot(mesh)
points(BlackSmokePrefData$east, BlackSmokePrefData$north, col = "red")
```

Now we define and fit our INLA model.

```{r model, results = 'hide', eval = FALSE}
spde_obj = inla.spde2.pcmatern(
  mesh = mesh,
  alpha = 2,
  prior.range = c(0.04, 0.05),
  prior.sigma = c(1, 0.01),
  constr = T
)
A_proj = inla.spde.make.A(
  mesh = mesh,
  loc = as.matrix(cbind(
    BlackSmokePrefData$east, BlackSmokePrefData$north
  )),
  group = BlackSmokePrefData$year - 65,
  # group membership needs to be 1:no_T
  n.group = no_T
)

s_index = inla.spde.make.index(name = "spatial.field",
                               n.spde = spde_obj$n.spde,
                               n.group = no_T)
time = (1:no_T) / no_T
time2 = time ^ 2

# create the stack object for estimating observation process y #

cov_y = data.frame(
  year = rep(time, each = no_sites),
  year_2 = rep(time2, each = no_sites),
  spatial_ind = rep(1:no_sites, times = no_T),
  spatial_ind2 = no_sites + rep(1:no_sites, times = no_T)
) # site-specific random intercepts

# Needs copies to include covariate twice

s_index_copy = s_index
names(s_index_copy) = c('spatial.field.copy',
                        "spatial.field.group.copy",
                        "spatial.field.repl.copy")

s_index_copy2 = s_index
names(s_index_copy2) = c('spatial.field.copy2',
                         "spatial.field.group.copy2",
                         "spatial.field.repl.copy2")


stack_y_est = inla.stack(
  data = list(
    y = BlackSmokePrefData$bsmoke,
    #single model
    alldata = cbind(BlackSmokePrefData$bsmoke, NA, NA),
    Ntrials = rep(0, times = length(BlackSmokePrefData$bsmoke))
  ),
  #joint model
  A = list(A_proj, A_proj, A_proj, 1),
  effects = list(
    c(s_index, list(Intercept = 1)),
    c(s_index_copy, list(Intercept_copy = 1)),
    c(s_index_copy2, list(Intercept_copy2 = 1)),
    cov_y
  ),
  tag = 'y_est'
)

formula_naive = y ~ -1 + Intercept +
  f(spatial.field, model = spde_obj) +
  f(spatial.field.copy, I(spatial.field.group.copy / no_T), model = spde_obj) +
  f(spatial.field.copy2, I((spatial.field.group.copy2 / no_T) ^ 2), model = spde_obj) +
  I(spatial.field.group / no_T) + I((spatial.field.group / no_T) ^ 2) +
  f(year, model = 'ar1') +
  f(spatial_ind,
    model = "iid2d",
    n = no_sites * 2,
    constr = TRUE) + #random site-specific intercepts
  f(spatial_ind2, year, copy = "spatial_ind")

theta.ini = c(
  1.597900,
  -1.277423,
  -0.443820,
  -1.441220,
  0.036510,
  -1.441336,
  0.016919,
  4.462918,
  1.437147,
  4,
  4,
  4
)
out.naive = inla(
  formula_naive,
  family = 'gaussian',
  data = inla.stack.data(stack_y_est),
  control.predictor = list(A = inla.stack.A(stack_y_est), compute = F),
  control.compute = list(dic = F, config = T, cpo = F),
  control.inla = list(strategy = "gaussian", int.strategy = 'eb'),
  control.mode = list(theta = theta.ini, restart = T),
  verbose = T,
  num.threads = 2
)
```

We can now take posteriors from the model.

```{r posterior, eval = FALSE}

m_samples = 50

samp <- inla.posterior.sample(m_samples, out.naive)

for (i in 1:m_samples)
{
samp[[i]]$latent = samp[[i]]$latent[-c(grep('Predictor', rownames(samp[[i]]$latent), fixed = T)), ]
}
```

Finally, summarize and plot the outputs:

```{r summarize, eval = FALSE}
stepsize = 5000 / sd_x
nxy = round(c(diff(range(
  BlackSmokePrefData$east
)),
diff(range(
  BlackSmokePrefData$north
))) / stepsize)  ## how many points?
proj_grid = inla.mesh.projector(
  mesh,
  xlim = range(BlackSmokePrefData$east),
  ylim = range(BlackSmokePrefData$north),
  dims = nxy
)
A.grid = inla.spde.make.A(mesh,
                          loc = cbind(BlackSmokePrefData$east, BlackSmokePrefData$north))

## define placeholders

tmp2 = matrix(0, nrow = no_T, ncol = mesh$n)
year = (1:no_T) / no_T
year2 = year ^ 2
Post_Array2 = array(0, dim = c(m_samples, no_T, mesh$n))
Post_Sites_Array2 = array(0, dim = c(m_samples, no_T, dim(BlackSmokePrefData)[1]))
residuals_array = array(0, dim = c(m_samples, no_T, dim(BlackSmokePrefData)[1]))
RI_array = array(0, dim = c(m_samples, no_T, dim(BlackSmokePrefData)[1])) # random intercepts
RS_array = array(0, dim = c(m_samples, no_T, dim(BlackSmokePrefData)[1])) # random slopes

for (i in 1:m_samples) {
  # Creating predictions for Y process
  for (j in 1:no_T)
    # loop over the years
  {
    Post_Array2[i, j, ] <- samp[[i]]$latent['Intercept:1'] + # Intercept
      samp[[i]]$latent[which(startsWith(prefix = 'year:', x = names(samp[[i]]$latent)))][j] * year[j] + # linear fixed effect
      #samp[[i]]$latent['year_2'] * year2[j] + # quadratic fixed effect
      as.numeric(samp[[i]]$latent[which(startsWith(prefix = 'spatial.field:', x = names(samp[[i]]$latent)))]) + # Spatial Intercept
      as.numeric(samp[[i]]$latent[which(startsWith(prefix = 'spatial.field.copy:', x = names(samp[[i]]$latent)))]) * year[j] + # Spatial linear slope
      as.numeric(samp[[i]]$latent[which(startsWith(prefix = 'spatial.field.copy2:', x = names(samp[[i]]$latent)))]) * year2[j]  # Spatial quadratic slope
    # Project onto the site locations and add random site-specific effects #
    Post_Sites_Array2[i, j, ] = as.numeric(A.grid %*% Post_Array2[i, j, ]) +
      as.numeric(samp[[i]]$latent[which(startsWith(prefix = 'spatial_ind:', x = names(samp[[i]]$latent)))[1:no_sites]]) + # random Intercept
      as.numeric(samp[[i]]$latent[which(startsWith(prefix = 'spatial_ind2:', x = names(samp[[i]]$latent)))[no_sites +
                                                                                                             (1:no_sites)]]) * year[j] # random slope
    # Extract the random intercepts #
    RI_array[i, j, ] = as.numeric(samp[[i]]$latent[which(startsWith(prefix = 'spatial_ind:', x = names(samp[[i]]$latent)))[1:no_sites]])
    
    # Extract the random slopes #
    RS_array[i, j, ] = as.numeric(samp[[i]]$latent[which(startsWith(prefix = 'spatial_ind2:', x = names(samp[[i]]$latent)))[no_sites +
                                                                                                                              (1:no_sites)]])
  }
}

## summarizing

Post_mean2 <-
  t(apply(Post_Array2, c(2, 3), function(x) {
    mean(x, na.rm = TRUE)
  }))
Post_sd2 <-
  t(apply(Post_Array2, c(2, 3), function(x) {
    sd(x, na.rm = TRUE)
  }))
Post_LCL2 <-
  apply((apply(Post_Array2, c(1, 2), function(x) {
    mean(x, na.rm = TRUE)
  })), 2, quantile, probs = c(0.025))
Post_UCL2 <-
  apply((apply(Post_Array2, c(1, 2), function(x) {
    mean(x, na.rm = TRUE)
  })), 2, quantile, probs = c(0.975))
RI_array <-
  t(apply(RI_array, c(2, 3), function(x) {
    mean(x, na.rm = TRUE)
  }))
RS_array <-
  t(apply(RS_array, c(2, 3), function(x) {
    mean(x, na.rm = TRUE)
  }))

## placeholders for uniform grid posterior blacksmoke values (not necessarily in dataset)

post_Matrix_grid_mean2 = array(NA, dim = c(nxy[1], nxy[2], no_T))
post_Matrix_grid_sd2 = array(NA, dim = c(nxy[1], nxy[2], no_T))
for (i in 1:no_T)
{
  post_Matrix_grid_mean2[, , i] = inla.mesh.project(proj_grid, Post_mean2[, i])
  post_Matrix_grid_sd2[, , i] = inla.mesh.project(proj_grid, Post_sd2[, i])
}

## placeholders for observed (R1 = active, R0 = inactive) site blacksmoke values

R1_results2 = matrix(NA, nrow = no_T, ncol = 4) # prediction means, sd, LCL, UCL
colnames(R1_results2) = c('predciction mean', 'prediction sd', 'LCL', 'UCL')
R0_results2 = matrix(NA, nrow = no_T, ncol = 4) # prediction means, sd, LCL, UCL
colnames(R0_results2) = c('predciction mean', 'prediction sd', 'LCL', 'UCL')
Grid_results2 = matrix(NA, nrow = no_T, ncol = 4) # prediction means, sd, LCL, UCL
colnames(Grid_results2) = c('predciction mean', 'prediction sd', 'LCL', 'UCL')

BlackSmokePrefData_original <-
  read.csv("data/BlackSmokePrefData.csv") ## need original columns here

for (j in 1:no_T)
{
  R1_results2[j, 1] = mean(Post_Sites_Array2[, j, which(!is.na(BlackSmokePrefData_original[, 3 +
                                                                                             j]))], na.rm = T)
  R1_results2[j, 2] = sd(apply(Post_Sites_Array2[, j, which(!is.na(BlackSmokePrefData_original[, 3 +
                                                                                                 j]))], c(1), mean, na.rm = T))
  R1_results2[j, 3] = quantile(apply(Post_Sites_Array2[, j, which(!is.na(BlackSmokePrefData_original[, 3 +
                                                                                                       j]))], 1, mean, na.rm = T), probs = 0.025)
  R1_results2[j, 4] = quantile(apply(Post_Sites_Array2[, j, which(!is.na(BlackSmokePrefData_original[, 3 +
                                                                                                       j]))], 1, mean, na.rm = T), probs = 0.975)
  
  R0_results2[j, 1] = mean(Post_Sites_Array2[, j, which(is.na(BlackSmokePrefData_original[, 3 +
                                                                                            j]))], na.rm = T)
  R0_results2[j, 2] = sd(apply(Post_Sites_Array2[, j, which(is.na(BlackSmokePrefData_original[, 3 +
                                                                                                j]))], c(1), mean, na.rm = T))
  R0_results2[j, 3] = quantile(apply(Post_Sites_Array2[, j, which(is.na(BlackSmokePrefData_original[, 3 +
                                                                                                      j]))], 1, mean, na.rm = T), probs = 0.025)
  R0_results2[j, 4] = quantile(apply(Post_Sites_Array2[, j, which(is.na(BlackSmokePrefData_original[, 3 +
                                                                                                      j]))], 1, mean, na.rm = T), probs = 0.975)
}

for (i in 1:no_T)
{
  Grid_results2[i, 1] = mean(post_Matrix_grid_mean2[, , i], na.rm = T)
  Grid_results2[i, 2] = mean(post_Matrix_grid_sd2[, , i], na.rm = T)
}
Grid_results2[, 3] = Post_LCL2
Grid_results2[, 4] = Post_UCL2

## Revert to the original scale (exponentiating and scaling up)
Grid_df_2 = data.frame(
  y = exp(c(
    Grid_results2[, 1], R1_results2[, 1], R0_results2[, 1]
  )) * mean(yearmeans),
  x = rep(66:96, times = 3),
  ymin = exp(c(Post_LCL2, R1_results2[, 3], R0_results2[, 3])) *
    mean(yearmeans),
  ymax = exp(c(Post_UCL2, R1_results2[, 4], R0_results2[, 4])) *
    mean(yearmeans),
  group = c(
    rep('Whole UK', times = 31),
    rep('R1', times = 31),
    rep('R0', times = 31)
  )
)

## plotting

Grid_plot_2 = ggplot(aes(
  x = x,
  y = y,
  ymin = ymin,
  ymax = ymax,
  alpha = 0.05
), data = Grid_df_2) +
  geom_ribbon(aes(
    colour = group,
    alpha = 0.05,
    fill = group
  )) +
  xlab('year') + ylab('posterior mean bs on transformed scale') +
  ggtitle(
    'Naive posterior means of BS at the selected, unselected sites and across the UK (R1, R0 and Whole UK resp.). '
  )
Grid_plot_2
```