---
output:
  pdf_document: default
  html_document: default
---
# Environmental hazards-spatial models {#hazards}

This chapter contains the basic theory for spatial processes and a number of approaches to modelling point-referenced spatial data. From this chapter, the reader will have gained an understanding of the following topics:

- Visualization techniques needed for both exploring and analyzing spatial data and communicating its features through the use of maps.
- Exploring the underlying structure of spatial data and methods for characterizing dependence over space.
- Second-order theory for spatial processes including the covariance. The variogram for measuring spatial associations.
- Stationarity and isotropy.
- Methods for spatial prediction, using both classical methods (kriging) as well as modern methods (Bayesian kriging).
- Non-stationarity fields.

<!-- NOTE: Unlike some of the other chapters you have to run this code sequentially -->


## Example 10.1 Spatial patterns of benzene concentrations in Montreal, QC, Canada {-}

Map the locations of the monitoring stations in Montreal.

```{r Ex 10.1 load data, message = FALSE, warning=FALSE, error=FALSE}
library(cowplot)
library(geoR)
library(ggmap)
library(spdep)
# Load data on benzene concentration in Montreal
benzene <- read.csv("data/montreal_benzene_apr.csv")
# TODO: Add description of the data, this is the April campaign
```

```{r Ex 10.1 map data, message = FALSE, warning=FALSE, error=FALSE}
# create a new variable in "sp" format and define coordinates
benzene_geo <- benzene
coordinates(benzene_geo) <- ~ lon + lat
proj4string(benzene_geo) <- CRS("+proj=longlat +datum=WGS84")
# specify the bounding box
latLongBox = bbox(benzene_geo)
location = c(latLongBox[1, 1] - 0.05,
             latLongBox[2, 1] - 0.05,
             latLongBox[1, 2] + 0.1,
             latLongBox[2, 2] + 0.05)
# create map with location dots marked on it in
MontrelBenzeneMap <-
  get_stamenmap(bbox =  location,
                zoom = 10)
ggmap(MontrelBenzeneMap) + geom_point(data = benzene,
                                      aes(x = lon,
                                          y = lat,
                                          size = Benzene), 
                                      col = "#011f4b",
                                      alpha = 0.45) + theme_void()

```

Using the `geoR` package we can also plot the following:

-  the locations of the sampling sites 
-  the concentrations of ozone in relation to the x and y coordinates
- and a histogram of the concentrations indicating the distribution of concentrations together with an estimate of the density.

```{r Ex 10.1 geoR plot}
# convert data to utm coordinates
benzene_utm <- spTransform(benzene_geo,
                        CRS("+proj=utm +zone=18 +ellps=WGS72"))
# Save the utm as a data frame
benzene_utm_df <- as.data.frame(benzene_utm)
colnames(benzene_utm_df) <- c("Benzene", "X", "Y")
# Save as geodata to generate the geodata plot
benzene_geodata <- as.geodata(benzene_utm_df,
                          coords.col = 2:3, data.col = 1)
plot(benzene_geodata)
```


## Example 10.2: Examining the log concentrations of benzene in Montreal {-}

```{r Ex 10.2 plot hist and qqplot log}
# Histogram for the log of the benzene concentration
par(mfrow=c(1,2))
log_histogram <- hist(log(benzene$Benzene), main = "", xlab = "log(Benzene)")
qqnorm(log(benzene$Benzene), bty = "n")
qqline(log(benzene$Benzene))

```
<!-- NOTE: I don't know how to generate the plot in Figure 9.4 -->

## Example 10.3: Mapping the locations of ozone monitoring sites in New York State {-}

```{r Ex 10.3 plot ozone sites NY, message = FALSE, warning=FALSE, error=FALSE}
# Load the metadata giving the site coordinates 
ny_data <- read.csv("data/NY_metadata.txt", sep="") 
# Now copy ny_data into ny_data_sp and convert data to "sp" format 
ny_data_sp <- ny_data
coordinates(ny_data_sp) <- ~Longitude+Latitude 

# assign a reference system to ny_data_sp 
proj4string(ny_data_sp) <- CRS("+proj=longlat +ellps=WGS84") 

# We next specify a bounding box - a 2 x 2 matrix of corners of the geographic 
# area. Then specify the range of locations within the box. 
# Note: location must bounding box format be in left -bottom -right -top 
latLongBox  <-  bbox(ny_data_sp)
location <-  c(latLongBox [1, 1] - 0.2 ,
               latLongBox [2, 1] - 0.2,
               latLongBox [1, 2] + 0.2 ,
               latLongBox [2, 2] + 0.2)

# Now create the map with location dots 
NYmap <-
  get_stamenmap(
   bbox = location,
   zoom = 8
  )

ggmap(NYmap) + geom_point(
  data = ny_data,
  aes(x = Longitude , y = Latitude),
  size = 4,
  color = "darkred"
) + theme_void()

```


## Example 10.5: Variogram {-}

```{r Ex 10.5 plot variogram}
library(gstat)
benzene_utm_geo <- benzene_utm_df
# get the coordinates in kms and turn into a Spatial object
benzene_utm_geo[,c("X", "Y")] <- benzene_utm_geo[,c("X", "Y")]/1000
coordinates(benzene_utm_geo) <- ~ X + Y
# Estimate variogram intercept only
benzene_inter_vgm <- gstat::variogram(log(Benzene)~ 1, data = benzene_utm_geo,
                         cutoff = 20, # cutoff distance
                         width = 20/10 # bins width 
                         )

benzene_inter_vgm_fit <- fit.variogram(benzene_inter_vgm,
                                model = vgm(0.1, "Exp", 15, 0.02))

benzene_inter_vgm_fit

# Estimate variogram using coordinates
benzene_vgm <- variogram(log(Benzene)~ X + Y, data = benzene_utm_geo,
                         cutoff = 20, # cutoff distance
                         width = 20/10 # bins width
                         )

benzene_vgm_fit <- fit.variogram(benzene_vgm,
                                model = vgm(0.1, "Exp", 3, 0.02))
benzene_vgm_fit

plot_inter_variog <- plot(benzene_inter_vgm, benzene_inter_vgm_fit, bty = "n")
plot_coord_variog <- plot(benzene_vgm, benzene_vgm_fit, bty = "n")

plot_grid(plot_inter_variog, plot_coord_variog, labels = "auto")
```

## Example 10.6 Basic spatial modelling and prediction of benzene in Montreal {-}


```{r Ex 10.6 kriging predictions, message = FALSE, warning=FALSE, error=FALSE}
# Generate grid
MtlPred <- expand.grid(seq (580, 615 , 0.5),
                             seq (5020, 5060 , 0.5) )
# change names grid
names(MtlPred)[ names(MtlPred)=="Var1"] <- "X"
names(MtlPred)[ names(MtlPred)=="Var2"] <- "Y"
# make the grid a Spatial object
coordinates (MtlPred) = ~ X + Y
gridded(MtlPred) = TRUE
# define the model based on the variogram fit from the previous example
mod <- vgm (0.053 , "Exp", 5.54, 0.0122)
# use ordinary kriging to predict values in the grid
x <- krige(log(Benzene) ~ X + Y, benzene_utm_geo , MtlPred , model = mod )
# Plot the ordinary kriging predictions and their variance

monitor_loc <- list('sp.points', benzene_utm_geo, pch=19, cex=.8, col='cornsilk4')

krig_pred <-
  spplot (
    x["var1.pred"],
    main = "Ordinary kriging predictions ",
    col.regions = viridis::plasma(60),
    sp.layout = list(monitor_loc),
    at = seq(-0.3, 0.8, 0.02)
  )
krig_var <-
  spplot (
    x["var1.var"],
    main = "Ordinary kriging variance ",
    col.regions = viridis::plasma(60),
    sp.layout = list(monitor_loc),
    at = seq(0, 0.2, 0.01)
  )

plot_grid(krig_pred, krig_var, labels = "auto")

```


## Example 10.9 Spatial modelling of Benzene in Montreal {-}

### Nimble {-}


```{r Ex 10.9 load data, message = FALSE, warning=FALSE, error=FALSE}
library(coda)
library(geoR)
library(magrittr)
library(nimble)
library(spdep)
library(tidyverse)
library(tidybayes)

```

```{r Ex 10.9 benzene to spatial data, message = FALSE, warning=FALSE, error=FALSE}
# Load data on benzene concentration in Montreal
benzene <- read.csv("data/montreal_benzene_apr.csv")
# create a new variable in "sp" format and define coordinates
benzene_geo <- benzene
coordinates(benzene_geo) <- ~ lon + lat
proj4string(benzene_geo) <- CRS("+proj=longlat +datum=WGS84")

benzene_utm <- spTransform(benzene_geo,
                        CRS("+proj=utm +zone=18 +ellps=WGS72"))
# Save the utm as a data frame
benzene_utm_df <- as.data.frame(benzene_utm)
# Change coordinates to kilometers and observations to the log scale
Mtl_benzene_sample <-
  data.frame(
    y = log(benzene_utm_df$Benzene),
    easting =  benzene_utm_df$lon / 1000,
    northing = benzene_utm_df$lat / 1000
  )
# Compute the distance matrix
obsCoords <- unname(as.matrix(Mtl_benzene_sample[,c("easting", "northing")]))

obsDist <- fields::rdist(obsCoords)
  
```


```{r Ex 10.9 nimble model}

Example10_9Code <-  nimbleCode ({
  # Covariance matrix spatial effect
  Sigma[1:n, 1:n] <-
    sigma_sq * exp(-distMatrix[1:n, 1:n] / phi) + tau_sq * identityMatrix(d = n)

    for (site in 1:n) {
    mean.site[site] <-
      beta0 + beta1 * easting[site] + beta2 * northing[site]
  }
  y[1:n]  ~  dmnorm(mean.site[1:n], cov = Sigma[1:n, 1:n])
  # Set up the priors for the spatial model
  sigma ~ T(dt(mu = 0, sigma = 1, df = 1), 0, Inf)
  sigma_sq <- sigma^2
  tau ~ T(dt(mu = 0, sigma = 1, df = 1), 0, Inf)
  tau_sq <- tau^2
  phi_inv ~ dgamma(shape =  5, rate = 5)
  phi <- 1 / phi_inv
  # prior for the coefficients
  beta0 ~ dnorm (0, 10)
  beta1 ~ dnorm (0, 10)
  beta2 ~ dnorm (0, 10)
  
}) 

```

Define the constants, data and initials lists for the `nimble` model.

```{r nimble Ex 10.9 set constants and inits, error = FALSE, message = FALSE, results='hide'}
# Define the constants, data, parameters and initial values
set.seed(1)
easting_scaled <- as.vector(scale(Mtl_benzene_sample$easting))
northing_scaled <- as.vector(scale(Mtl_benzene_sample$northing))

constants <-
  list(n = nrow(Mtl_benzene_sample))
ex.data <-
  list(y = Mtl_benzene_sample$y,
       easting = easting_scaled,
       northing = northing_scaled,
       distMatrix = obsDist)
params <- c( "beta0",  "beta1","beta2", "phi",  "tau", "sigma", "tau_sq", "sigma_sq")
inits <- list( sigma = 0.1, phi_inv = 6/max(obsDist), tau = 0.1)
# Run model in nimble
start_time <- Sys.time()

mcmc.out <- nimbleMCMC(
  code = Example10_9Code,
  constants = constants,
  data = ex.data,
  inits = inits,
  monitors = params,
  niter = 40000,
  nburnin = 20000,
  thin = 14,
  WAIC = TRUE,
  nchains = 2,
  summary = TRUE,
  samplesAsCodaMCMC = TRUE
)
end_time <- Sys.time()
run_time <- end_time - start_time
run_time

```

```{r Ex 10.9 nimble WAIC, ESS and traceplots}
mcmc.out$WAIC
min(coda::effectiveSize(mcmc.out$samples))
plot(mcmc.out$samples[, c("beta0")], bty = "n", main = "beta0")
plot(mcmc.out$samples[, c("beta1")], bty = "n", main = "beta1")
plot(mcmc.out$samples[, c("beta2")], bty = "n", main = "beta2")
plot(mcmc.out$samples[, c("sigma")], bty = "n", main = "sigma")
plot(mcmc.out$samples[, c("tau")], bty = "n", main = "tau")
plot(mcmc.out$samples[, c("phi")], bty = "n", main = "phi")
mcmc.out$summary$all.chains[c("beta0", "beta1", "beta2", "sigma_sq", "tau_sq", "phi"), ]
```

```{r Ex 10.9 nimble rediction locations}
# Obtain coordinates for predictions 
# note that we are using the same coordinates as the one generated for 
# the kriging example
Mtl_centroids_df <- as.data.frame(MtlPred)
predCoords <- unname(as.matrix(Mtl_centroids_df))

```

Following the posterior predictive distribution we have to define a model for the predictions. 

```{r Ex 10.9 nimble predictions}
# Extract samples from nimble model
tidy_post_samples <- mcmc.out$samples %>% tidy_draws()

# Extract posterior samples for each of the parameters of interest
post_beta0 <- tidy_post_samples$beta0
post_beta1 <- tidy_post_samples$beta1
post_beta2 <- tidy_post_samples$beta2
post_sigmasq <- tidy_post_samples$sigma_sq
post_phi <- tidy_post_samples$phi
post_tausq <- tidy_post_samples$tau_sq

# Scale coordinates for predictive locations
predCoords_sc <- predCoords 
predCoords_sc[,1] <-
  (predCoords_sc[,1] - mean(Mtl_benzene_sample$easting)) / sd(Mtl_benzene_sample$easting)
predCoords_sc[,2] <- 
  (predCoords_sc[,2] - mean(Mtl_benzene_sample$northing)) / sd(Mtl_benzene_sample$northing)
  
n0 <- nrow(predCoords_sc)
L <- length(post_tausq)

obsMu <- cbind(1, easting_scaled, northing_scaled) %*% t(cbind(post_beta0, post_beta1, post_beta2))

predMu <- cbind(1, as.matrix(predCoords_sc)) %*% t(cbind(post_beta0, post_beta1, post_beta2))

pred2obsDist <- fields::rdist(predCoords, obsCoords)

```


```{r Ex 10.9 nimble predictions using Rcpp}

Rcpp::sourceCpp("functions/prediction_marginal_gp12.cpp")
args(prediction_marginal_gp12)
system.time(
  pred_samples <- prediction_marginal_gp12(
    y = Mtl_benzene_sample$y,
    obsMu = obsMu,
    predMu = predMu,
    obsDistMat = obsDist,
    pred2ObsDistMat = pred2obsDist,
    sigmasq = post_sigmasq,
    phi = post_phi,
    tausq = post_tausq,
    iterprint = 1000
  )
)

str(pred_samples)

predict_res_dt <- data.frame(
  xcoord = predCoords[, 1],
  ycoord = predCoords[, 2],
  post.mean = apply(pred_samples, 1, mean),
  post.var = apply(pred_samples, 1, var),
  q2.5 = apply(pred_samples, 1, function(x)
    quantile(x, prob = 0.025)),
  q50 = apply(pred_samples, 1, function(x)
    quantile(x, prob = 0.5)),
  q97.5 = apply(pred_samples, 1, function(x)
    quantile(x, prob = 0.975))
)


```



```{r Ex 10.9 nimble plot predictions, messages = FALSE}

x$nimble.pred <- predict_res_dt$post.mean
x$nimble.var <- predict_res_dt$post.var

nimble_pred <-
  spplot (
    x["nimble.pred"],
    main = "Nimble spatial predictions ",
    col.regions = viridis::plasma(60),
    sp.layout = list(monitor_loc),
    at = seq(-0.3, 0.8, 0.02)
  )
nimble_var <-
  spplot (
    x["nimble.var"],
    main = "Nimble predictions variance ",
    col.regions = viridis::plasma(60),
    sp.layout = list(monitor_loc),
    at = seq(0, 0.2, 0.01)

  )

plot_grid(nimble_pred, nimble_var, labels = "auto")

```

### Stan {-}

```{r load stan, message=FALSE, echo = FALSE, warning=FALSE, message= FALSE}
# load rstan with options
library(geoR)
library(rstan)
library(spdep)
options(mc.cores = parallel::detectCores())
rstan_options(auto_write = TRUE)
```


```{r stan load data}
benzene_data <- read.csv("data/montreal_benzene_apr.csv")
# create a new variable in "sp" format and define coordinates
benzene_data_geo <- benzene_data
coordinates(benzene_data_geo) <- ~ lon + lat
proj4string(benzene_data_geo) <- CRS("+proj=longlat +datum=WGS84")

benzene_data_utm <- spTransform(benzene_data_geo,
                        CRS("+proj=utm +zone=18 +ellps=WGS72"))
# Save the utm as a data frame
benzene_utm_df <- as.data.frame(benzene_data_utm)
# change the observed values to the log scale and the coordinates to km's
Mtl_benzene_sample <-
  data.frame(
    y = log(benzene_utm_df$Benzene),
    easting =  benzene_utm_df$lon / 1000,
    northing = benzene_utm_df$lat / 1000
  )
# Compute the distance matrix
obsCoords <- unname(as.matrix(Mtl_benzene_sample[,c("easting", "northing")]))

obsDist <- fields::rdist(obsCoords)

```



```{r engine='bash', comment='', echo = FALSE}
cat functions/Example10_9.stan
``` 


```{r Ex 10.9 stan exponential model, warning = FALSE, message = FALSE}
set.seed(1)
easting_scaled <- as.vector(scale(Mtl_benzene_sample$easting))
northing_scaled <- as.vector(scale(Mtl_benzene_sample$northing))

N <- nrow(benzene_utm_df)
# Change coordinates to kilometers
X <- data.frame(easting = easting_scaled,
                northing = northing_scaled) 

ex.data <- list(
  N = N,
  p = 2,
  y =   log(benzene_utm_df$Benzene),
  dist_matrix = obsDist,
  X = as.matrix(X)
)

Example10_9Stan  <- stan(
  file = "functions/Example10_9.stan",
  data = ex.data,
  warmup = 10000,
  iter = 20000,
  chains = 2,
  thin = 10,
  pars = c("beta0", "beta", "sigma_sq", "tau_sq", "phi"),
  include = TRUE
)

```


```{r  Ex 10.9 stan exp traceplots, warning = FALSE}

rstan::traceplot(Example10_9Stan, pars = c("beta0","beta","sigma_sq", "tau_sq", "phi"))

```

```{r Ex 10.9 stan exp results}
summary_exp_stan <-
  summary(
    Example10_9Stan,
    pars = c("beta0","beta", "phi", "sigma_sq", "tau_sq"),
    probs = c(0.025, 0.975)
  )

summary_exp_stan$summary

```


```{r Ex 10.9 stan rediction locations}
# Obtain coordinates for predictions 
# note that we are using the same coordinates as the one generated for 
# the kriging example
Mtl_centroids_df <- as.data.frame(MtlPred)
predCoords <- unname(as.matrix(Mtl_centroids_df))

```

```{r Ex 10.9 stan predictions}
# Extract samples from nimble model
stan_post_samples <- rstan::extract(Example10_9Stan,
    pars = c("beta0","beta", "phi", "sigma_sq", "tau_sq"))

# Extract posterior samples for each of the parameters of interest
post_beta0 <- stan_post_samples$beta0
post_beta <- stan_post_samples$beta
# post_beta2 <- stan_post_samples$beta2
post_sigmasq <- stan_post_samples$sigma_sq
post_phi <- stan_post_samples$phi
post_tausq <- stan_post_samples$tau_sq

# Scale coordinates for predictive locations
predCoords_sc <- predCoords 
predCoords_sc[,1] <-
  (predCoords_sc[,1] - mean(Mtl_benzene_sample$easting)) / sd(Mtl_benzene_sample$easting)
predCoords_sc[,2] <- 
  (predCoords_sc[,2] - mean(Mtl_benzene_sample$northing)) / sd(Mtl_benzene_sample$northing)
  
n0 <- nrow(predCoords_sc)
L <- length(post_tausq)

obsMu <- cbind(1, easting_scaled, northing_scaled) %*% t(cbind(post_beta0, post_beta))

predMu <- cbind(1, as.matrix(predCoords_sc)) %*% t(cbind(post_beta0, post_beta))

pred2obsDist <- fields::rdist(predCoords, obsCoords)

```


```{r Ex 10.9 stan predictions using Rcpp}

Rcpp::sourceCpp("functions/prediction_marginal_gp12.cpp")
args(prediction_marginal_gp12)
system.time(
  pred_samples <- prediction_marginal_gp12(
    y = Mtl_benzene_sample$y,
    obsMu = obsMu,
    predMu = predMu,
    obsDistMat = obsDist,
    pred2ObsDistMat = pred2obsDist,
    sigmasq = post_sigmasq,
    phi = post_phi,
    tausq = post_tausq,
    iterprint = 1000
  )
)

predict_res_dt <- data.frame(
  xcoord = predCoords[, 1],
  ycoord = predCoords[, 2],
  post.mean = apply(pred_samples, 1, mean),
  post.var = apply(pred_samples, 1, var),
  q2.5 = apply(pred_samples, 1, function(x)
    quantile(x, prob = 0.025)),
  q50 = apply(pred_samples, 1, function(x)
    quantile(x, prob = 0.5)),
  q97.5 = apply(pred_samples, 1, function(x)
    quantile(x, prob = 0.975))
)


```



```{r Ex 10.9 stan plot predictions, messages = FALSE}

x$stan.pred <- predict_res_dt$post.mean
x$stan.var <- predict_res_dt$post.var


stan_pred <-
  spplot (
    x["stan.pred"],
    main = "Stan spatial predictions ",
    col.regions = viridis::plasma(60),
    sp.layout = list(monitor_loc),
    at = seq(-0.3, 0.8, 0.02)
  )
stan_var <-
  spplot (
    x["stan.var"],
    main = "Stan predictions variance ",
    col.regions = viridis::plasma(60),
    sp.layout = list(monitor_loc),
    at = seq(0, 0.2, 0.01)
  )

plot_grid(stan_pred, stan_var, labels = "auto")

```

## Example 10.11: INLA, creating a mesh {-}

```{r Ex 10.11 load data for INLA, message = FALSE, warning=FALSE, error=FALSE}
library(geoR)
library(INLA)
library(spdep)

benzene_data <- read.csv("data/montreal_benzene_apr.csv")
# create a new variable in "sp" format and define coordinates
benzene_data_loc <- benzene_data
coordinates(benzene_data_loc) <- ~ lon + lat
proj4string(benzene_data_loc) <- CRS("+proj=longlat +datum=WGS84")

benzene_data_utm <- spTransform(benzene_data_loc,
                        CRS("+proj=utm +zone=18 +ellps=WGS72"))
# Save the utm as a data frame
locations_df <- as.data.frame(benzene_data_utm@coords)

benzene_utm_df <- data.frame( ID = 1:nrow(benzene_data),
                              X = benzene_data_utm@coords[,1]/1000,
                              Y = benzene_data_utm@coords[,2]/1000,
                              logbenzene = log(benzene_data_utm$Benzene))
  
# change the observed values to the log scale and the coordinates to km's

mesh = inla.mesh.create(locations_df, cutoff = 0.01, 
                       refine =(list(min.angle =20)))
plot(mesh , col="gray", main="") 


```

## Example 10.12: Fitting an SPDE model using R–INLA: benzene concentration in Montreal {-}

```{r Ex 10.12 spde inla, warning=FALSE}
# Field std . dev . for theta =0
sigma0 = 1
# find the range of the location data
size = min(c(diff(range(mesh$loc[, 1])),
             diff (range(mesh$loc[, 2]))))
# A fifth of the approximate domain width .
range0 = size/5
kappa0 = sqrt(8)/range0
tau0 = 1/(sqrt (4*pi)*kappa0*sigma0)
spde = inla.spde2.matern (
  mesh,
  B.tau = cbind(log (tau0), -1, +1),
  B.kappa = cbind(log (kappa0), 0, -1),
  theta.prior.mean = c(0 , 0),
  constr = TRUE
)

formula = logbenzene  ~ 1 + X + Y + f(ID , model = spde)
model = inla(
  formula,
  family = "gaussian",
  data = benzene_utm_df ,
  control.predictor = list(compute = TRUE),
  control.compute = list(dic = TRUE , config = TRUE)
)

model$summary.fixed
```

## Example 10.13: Directional variograms {-}

```{r Ex 10.13 directional variogram, error = FALSE, message = FALSE, results='hide'}
### Compute and plot the directional variogram
CA.geo <- as.geodata( benzene_utm_df , coords.col = 2:3 , data.col=1)
CA.vario4 <- variog4(CA.geo )
plot(CA.vario4)
```


<!-- REVIEW: I am not sure where this example goes -->

## Example 10.14: Spatial modeling of malaria in Gambia {-}

```{r Ex 10.14 clean, include = FALSE}
rm(list=ls())
```


Note: There might be a warning when loading the $\texttt{raster}$ package, if necessary, uninstall and re-install the package. 

For this example we are using the same $\texttt{gambia}$ data set from the $\texttt{geoR}$ package but an $\texttt{id_area}$ column was added using QGIS to differentiate the different areas as in the original paper.

```{r Ex 10.14 load libraries and data, message = FALSE, warning=FALSE, error=FALSE}

library(dplyr) # to manipulate the data
library(geoR) # to get the dataset
library(ggmap) # to plot the map
library(nimble) # for modeling
#library(raster) # to get the environmental data
library(rgdal) # for adding and transforming coordinates
library(sf) # manipulate spatial data
library(sp) # for manipulating spatial data
library(stringr) # to analyze posterior
library(viridis) # for a more cheerful color palette

# Note, we are using the same Gambia dataset as in the geoR package but an id_area
# attribute has been added using QGIS to a
gambia <- read.csv("data/gambia_area.csv") # gambia dataset from geoR package

theme_clear <- function(){
  theme(
    panel.grid.major = element_blank(), # remove background grid
    panel.grid.minor = element_blank(),
    panel.background = element_blank(), # remove grey background
    axis.line = element_line(colour = "black"), # keep axis line
    text=element_text(size=12)
  )
}
```

Since the data is given at the individual level, we want to aggregate the malaria tests by village. 
If we explore the data frame we see that there are 2035 individuals at 65 villages. 

```{r Ex 10.14 explore data}
head(gambia)
dim(gambia)
dim(unique(gambia[, c("x", "y")]))
```

We create a new data frame aggregated by village containing the coordinates, the number of malaria tests, and the prevalence. 

```{r Ex 10.14 aggregate data}

malaria_village <- group_by(gambia, x, y) |>
  summarize(total = n(),
            positive = sum(pos),
            prev = positive / total) |> as.data.frame()

head(malaria_village)
```

```{r Ex 10.14 plot prevalence, message = FALSE, warning=FALSE}

# create a new variable in "sp" format and define coordinates
malaria_utm <- malaria_village
coordinates(malaria_utm) <- ~ x + y
proj4string(malaria_utm) <- CRS("+proj=utm +zone=28")
# convert to long lat 
malaria_geo <- spTransform(malaria_utm, CRS("+proj=longlat +datum=WGS84"))
# add long lat coordinates to malaria dataframe
malaria_village[, c("long", "lat")] <- coordinates(malaria_geo)

# specify the bounding box
latLongBox = bbox(malaria_geo)
location = c(latLongBox[1, 1] - 0.05,
             latLongBox[2, 1] - 0.05,
             latLongBox[1, 2] + 0.05,
             latLongBox[2, 2] + 0.05)

# create map with location dots marked on it in
GambiaMap <-
  get_stamenmap(bbox =  location,
                zoom = 11,
                type = terrain-background)

ggmap(GambiaMap) + geom_point(data = malaria_village,
                              aes(x = long,
                                  y = lat,
                                  col = prev),
                              size = 2) + scale_color_viridis() + theme_void()

```

<!-- NOTE: I am not doing a Matern cause I don't know how to code it, I'll use an exponential instead -->




###  Nimble {-} 

```{r Ex 10.14 nimble model}

Example10_14_Nimble <- nimbleCode({
  
  # Define priors
  sigma ~ T(dt(mu = 0, sigma = 1, df = 1), 0, Inf)
  sigma_sq <- sigma ^ 2
  phi_inv ~ dgamma(shape =  5, rate = 5)
  phi <- 1 / phi_inv

  Sigma[1:N, 1:N] <-
     sigma_sq*(1 + (sqrt(3)*obs_dist_mat[1:N, 1:N])/phi) * exp(-sqrt(3)*obs_dist_mat[1:N, 1:N] / phi) 

   #mean_S[1:N] <- b0*ones[1:N]
  S[1:N] ~ dmnorm(zeroes[1:N], cov = Sigma[1:N,1:N])
  # 
    for (j in 1:n) {  # by child
    logit(p[j])  <-  b0 + inprod(b[1:k], X[j,1:k]) + S[index_village[j]]
    y[j] ~ dbern(p[j])
    }

  for (l in 1:k){
    b[l] ~ dnorm(0, var = 10)
  }
  
  b0 ~  dnorm(0, sd = 5)

})

```

```{r Ex 10.14 prepare data nimble model, warning = FALSE, error = FALSE, message = FALSE, results='hide'}
# distance specification
coords_sf <- sf::st_as_sf(malaria_village[,c("long","lat")], 
                          coords = c("long","lat")) |> 
  sf::st_set_crs(4326)
obs_dist_mat <- sf::st_distance(coords_sf)
obs_dist_mat <- units::set_units(obs_dist_mat, km)
obs_dist_mat <- units::set_units(obs_dist_mat, NULL)

# define indicator variables for each village

gambia_df <- mutate(
  gambia,
  id_child = 1:nrow(gambia), # add an id for each child
  value = 1, # this will be needed later for the villages
  id_village = as.numeric(interaction( # add an id for the villages
    x, y, drop = TRUE, lex.order = TRUE
  ))
)  |> 
  tidyr::spread(id_area, value, fill = 0) |>
  rename(area1 = '1', area2 = '2', area3 = '3', area4 = '4', area5 = '5')

## Model specification
# Variables matrix
X <- data.frame(age = scale(gambia_df[,"age"], center = TRUE, scale = FALSE),
                netuse = gambia_df[,"netuse"],
                treated = gambia_df[,"treated"],
                green = scale(gambia_df[,"green"], center = TRUE, scale = FALSE),
                phc = gambia_df[,"phc"],
                area2 = gambia_df[,"area2"],
                area3 = gambia_df[,"area3"],
                area4 = gambia_df[,"area4"],
                area5 = gambia_df[,"area5"]
                )

index_village <- gambia_df[,"id_village"]

n <- nrow(X) # child number
N <- nrow(malaria_village) # number of villages
zeroes <- rep(0, N) # auxiliary vector of zeroes for model
ones <- rep(1, N)

```



```{r Ex 10.14 run nimble}

const_list <- list(n = n, # number of childs
                   N = N, # number of villages
                   zeroes = zeroes, # vector of zeroes 
                   prior_max_dist = max(obs_dist_mat)/6, # max dist for phi prior
                   k =  ncol(X),# number of predictors
                   index_village = index_village) 

dat_list <- list(y = gambia_df$pos, # malaria positive test 
                 obs_dist_mat = obs_dist_mat, # distance matrix in km
                 X = X # predictors matrix
                 )

init_list <- list(sigma = 0.5, 
                  p = rep(expit(rnorm(1, 0, 1)), n), 
                  phi_inv = 6/max(obs_dist_mat),
                  b = rep(0, ncol(X)), 
                  b0 = 0,
                  S = rnorm(N, 0, 1))

#init_list <- list(p = runif(n, 0, 1), b = rnorm(ncol(X), 0,  1), b0 = rnorm(1, 0, 1))

Rmodel <-
  nimbleModel(
    Example10_14_Nimble,
    constants = const_list,
    data = dat_list,
    inits = init_list
  )
Rmodel$initializeInfo()
Cmodel <- compileNimble(Rmodel, showCompilerOutput = FALSE)
conf <-
  configureMCMC(Rmodel, monitors = c( "b0", "b", "p", "S", "sigma", "phi"))

# conf$removeSamplers(c('S'))
# conf$addSampler(target = c('S'), type = 'AF_slice')

Rmcmc <- buildMCMC(conf)
Cmcmc <-
  compileNimble(
    Rmcmc,
    project = Cmodel,
    resetFunctions = TRUE,
    showCompilerOutput = TRUE
  )
niters <- 30000
nburnins <- 0.5 * niters
nchains <- 2
nthins <- 14
post_samples <- runMCMC(
  Cmcmc,
  niter = niters,
  nburnin = nburnins,
  thin = nthins,
  nchains = nchains,
  samplesAsCodaMCMC = TRUE,
  summary = TRUE
)
```



```{r Ex 10.14 nimble traceplots ESS }

plot(post_samples$samples[, c("b0")], bty = "n", main = "b0")
plot(post_samples$samples[, c("b[1]")], bty = "n", main = "b[1]")
plot(post_samples$samples[, c("b[2]")], bty = "n", main = "b[2]")
plot(post_samples$samples[, c("b[3]")], bty = "n", main = "b[3]")
plot(post_samples$samples[, c("b[4]")], bty = "n", main = "b[4]")
plot(post_samples$samples[, c("b[5]")], bty = "n", main = "b[5]")
plot(post_samples$samples[, c("b[6]")], bty = "n", main = "b[6]")
plot(post_samples$samples[, c("sigma")], bty = "n", main = "sigma")
plot(post_samples$samples[, c("phi")], bty = "n", main = "phi")
plot(post_samples$samples[, c("S[1]")], bty = "n", main = "S[1]")
plot(post_samples$samples[, c("S[45]")], bty = "n", main = "S[45]")
plot(post_samples$samples[, c("S[54]")], bty = "n", main = "S[54]")
plot(post_samples$samples[, c("p[1]")], bty = "n", main = "p[1]")
plot(post_samples$samples[, c("p[1805]")], bty = "n", main = "p[1805]")

# Get minimum effective size (ESS) and which variable has the min ESS
min(coda::effectiveSize(post_samples$samples))
mcmc_variable_names <- colnames(post_samples$samples$chain1)

mcmc_variable_names[which(coda::effectiveSize(post_samples$samples) == min(coda::effectiveSize(post_samples$samples)))]

```



```{r Ex 10.14 nimble summary, echo = TRUE}

# Extract samples
variables <- c("b0", "b[1]", "b[2]", "b[3]", "b[4]", "b[5]", "b[6]", "b[7]", "b[8]", "b[9]" ,"sigma", "phi")
summary_nimble <- post_samples$summary$all.chains
summary_nimble[variables,]

```



```{r Ex 10.14 nimble summary p, echo = TRUE}
# Plot posterior summary for the spatial random effect by village
post_summary <- post_samples$summary$all.chains

post_sum_S <-
  as.data.frame(post_summary) |>  tibble::rownames_to_column() |>
  filter(str_detect(rowname, "S")) |>
  dplyr::select(rowname, `95%CI_low`, Mean, `95%CI_upp`)  |>
  mutate(village = gsub(".*?([0-9]+).*", "\\1", rowname))

post_sum_S$village <-
  factor(post_sum_S$village , levels = 1:65)

ggplot(data = post_sum_S, aes(x = village)) +
  geom_pointrange(aes(ymin = `95%CI_low`, ymax = `95%CI_upp`, y = Mean)) +
  geom_hline(yintercept = 0, linetype = "dotted") +
  scale_x_discrete(
    breaks = post_sum_S$village[seq(1, length(post_sum_S$village), by = 5)]) +
  theme_clear() +  ylab("") + xlab("village") +
  ggtitle("Posterior summary spatial random effect by village")

```




### Stan {-}

```{r Ex 10.14 load stan, message=FALSE, echo = FALSE, warning=FALSE, message= FALSE}
# load rstan with options
library(rstan)
options(mc.cores = parallel::detectCores())
rstan_options(auto_write = TRUE)
```


```{r engine='bash', comment='', echo = FALSE}
cat functions/Example10_14.stan
```


```{r Ex 10.14 stan data, warning = FALSE}

ex.data <- list( n = nrow(gambia), # number of children
                 k = ncol(X), # number of covariates
                 N = N, # number of villages
                 y = gambia$pos, # positive tests
                 dist_matrix = obs_dist_mat, # distance matrix in km
                 X = X, # altitude per village
                 index_village = index_village
                 )

Example10_14Stan  <- stan(
  file = "functions/Example10_14.stan",
  data = ex.data,
  warmup = 15000,
  iter = 30000,
  chains = 2,
  thin = 10,
  pars = c("beta0", "betas","sigma", "phi", "S"),
  include = TRUE
)

```

```{r Ex 10.14 stan traceplots}
#computing WAIC using the package loo

rstan::traceplot(Example10_14Stan, pars = c("beta0","betas","sigma",  "phi"))

```


```{r Ex 10.14 stan summary, echo = TRUE}

# Extract samples
summary_stan <-
  summary(
    Example10_14Stan,
    pars = c("beta0","betas", "sigma", "phi"),
    probs = c(0.025, 0.975)
  )

summary_stan$summary

```



```{r Ex 10.14 stan posterior plots}
S_summary <-
  summary(Example10_14Stan, pars = c("S"))$summary

S_summary_df <- data.frame(S_summary) |>
  tibble::rownames_to_column() |>
  filter(rowname %in% paste0("S[", 1:65, "]")) |>
  mutate(village = 1:65) |>
  dplyr::select(mean,  X2.5., X97.5., village)

S_summary_df$village <-
  factor(S_summary_df$village , levels = 1:65)


  ggplot(S_summary_df, aes(x = village, group = 1)) +
  geom_pointrange(aes(ymin = X2.5., ymax = X97.5., y = mean)) +
  geom_hline(yintercept = 0, linetype = "dotted") +
  scale_x_discrete(
    breaks = S_summary_df$village[seq(1, length(S_summary_df$village), by = 5)]) +
  theme_clear() +  ylab("") + xlab("village") +
  ggtitle("Posterior summary spatial random effect by village")



```

