# Time-why it also matters {#Time}

The chapter contains the theory required for handling time series data. From this chapter, the reader will have gained an understanding of the following topics:

- That a temporal process consists of both low and high frequency components, the former playing a key role in determining long-term trends while the latter may be associated with shorter-term changes.
- Techniques for the exploratory analysis of the data generated by the temporal process, including the ACF (correlogram) and PACF (periodogram).
- Models for irregular (high frequency) components after the regular components (trend) have been removed.
- Methods for forecasting, including exponential smoothing and ARIMA modelling.
- The state space modelling approach, which sits naturally within a Bayesian setting and which provides a general framework for most of the classical time series models and many more besides.
- Implementing time series processes within a Bayesian hierarchical framework.


## Example 11.1 Ground level ozone concentrations


```{r Ex 11.1 load data, message=FALSE, echo = TRUE, warning=FALSE, message= FALSE}

library(ggplot2)

# Load data
# Select one site from the daily data
site_daily <- read.csv("data/LA_ozone_daily.csv", header = TRUE)
site_daily$date <- as.Date(site_daily$date)

# load hourly data from that same site
site_hourly <- read.csv("data/LA_ozone_hourly.csv", header = TRUE)

```




```{r Ex 11.1 plot time series}

ggplot(data = site_daily,) +
  geom_line(aes(x = date, y = ozone, group = 1)) +
  xlab("Day in 2013 at Los Angeles Site 060379033") + 
  ylab("Max Daily 8hr Ozone Level (ppm)") +
  theme(
    panel.grid.major = element_blank(), # remove background grid
    panel.grid.minor = element_blank(),
    panel.background = element_blank(), # remove grey background
    axis.line = element_line(colour = "black"), # keep axis line
    axis.text.x = element_text(angle = 60, hjust = 1) # tilt x axis labels 60 deg
  )



ts.plot(site_hourly$value,
        ylab = "Hourly Ozone Concentration (ppm)",
        xlab = "Hour in 2013's ozone season")
abline(h = 0.075, v = 0, col = "gray60")
```





